An OpenCL-based tridiagonal solver for evaluation of compact finite differences on parallel architectures

### 2015-11-18 thesis/compact-finite-difference.tex
remove space after cite

Here, we consider the case of multiple GPUs
on a single shared memory node, i.e.,
multiple GPUs attached to the same PCI-e bus.
In this case, every GPU is visible to the host,
and the GPUs read from and write into
the same host memory space.
The domain is divided into a number of
into ``subdomains'',
as shown in \ref{fig:compact-shared-gpu}.
The domain decomposition is done such that
only a single subdomain is used along the
coordinate direction of the derivatives.
This is the method presented by
Sakharnykh et al. ~\cite{sakharnykhADIconf},
and it has the advantage that
no coordination between GPUs is required:
each GPU is assigned an independent
set of grid lines to solve.

>>>>

Here, we consider the case of multiple GPUs
on a single shared memory node, i.e.,
multiple GPUs attached to the same PCI-e bus.
In this case, every GPU is visible to the host,
and the GPUs read from and write into
the same host memory space.
The domain is divided into a number of
into ``subdomains'',
as shown in \ref{fig:compact-shared-gpu}.
The domain decomposition is done such that
only a single subdomain is used along the
coordinate direction of the derivatives.
This is the method presented by
Sakharnykh et al. \cite{sakharnykhADIconf},
and it has the advantage that
no coordination between GPUs is required:
each GPU is assigned an independent
set of grid lines to solve.

### 2015-11-18 thesis/introduction-and-background.tex
remove space after cite

    \item When the derivative is expressed as
        some combination of the function values
        on \emph{both} sides of $i$, i.e,
        at $\hdots$, $i-2$, $i-1$, $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{central} difference.
\end{enumerate}
%
The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy~\cite{fornberg1988generation}.
In the evaluation of spatial derivatives,
the above schemes are referred to as \emph{explicit}
schemes, as the derivative at each point
can be expressed as some explicit combination of function values.

@@

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{h} + \
    b\frac{f_{i+2} - f_{i-2}}{h} + \\
    c\frac{f_{i+3} - f_{i-3}}{h} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints ~\cite{kennedy1994several,lele1992compact},
and $h$ is the space between two grid points.
We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
and $A$ is, in general, a \emph{banded} matrix.

@@

The ability of a finite difference scheme
to accommodate large wavenumbers
is extremely relevant in
computational fluid dynamics applications
~\cite{KravchenkoEffNumErr}.
In DNS, the scheme
\emph{must} capture the
rapidly varying characteristics of the flow
associated with the turbulence.
For this purpose,
higher order explicit schemes
may be considered.
While they are straightforward to compute,
they are associated with large stencil widths.
In most applications,
large stencil widths are undesirable.
This is because
the arithmetic intensity increases with stencil size, i.e.,
a larger number of computations must be performed per grid point.
Further,
the amount of boundary information that must be exchanged
between parallel processes increases,
which can seriously affect overall performance.

@@

One of the advantages of the compact finite difference approach
is that it accommodates \emph{non-periodic} boundary conditions.
This is in contrast to other methods used in DNS,
such as \emph{spectral} methods.
We note that the Eq. (\ref{eqn:general-compact})
cannot be applied near the boundary points.
At the boundaries,
\emph{non-centered} or \emph{one-sided}
finite difference approximations are required.
Some considerations are made in choosing
these approximations:
firstly, the bandwidth of the resulting banded matrix
must be preserved.
Secondly,
the width of the boundary stencils
must be lower than the interior stencils,
as higher order boundary stencils are unstable
~\cite{kennedy1994several}.
In general, boundary equations for the first derivative
are of the following form ~\cite{lele1992compact}:

@@

Similar tridiagonal schemes are available
to evaluate higher order derivatives.
Compact schemes for
second, third and fourth derivatives
are derived in ~\cite{lele1992compact}.

@@

The evolution of graphics processing units (GPUs)
has traditionally been guided by the requirements
of graphics applications such as video gaming.
In general, these graphics computations involve
manipulations and calculations with pixel data.
Because these calculations are largely independent,
the GPU hardware is specialized
to perform several of them \emph{in parallel}.
Therefore, the
GPU may be regarded as a highly multithreaded processor.
Initially,
programming GPUs for scientific computation tasks
was a cumbersome process,
as the programming interfaces were designed for graphics computations,
and attempts to use them for other purposes were essentially ``hacks''.
Moreover, there was little support for double precision arithmetic,
which is critical for many applications.
But with the introduction of the
NVIDIA Compute Unified Device Architecture (CUDA)
parallel computing platform
and programming model,
CUDA-enabled GPUs have come into the
mainstream in high-performance computing.
A history of the development of GPUs
for general purpose computation,
and of CUDA specifically is available
in ~\cite{HwuBook}.
We provide the essential and relevant details about CUDA here,
and refer to the
CUDA Programming Guide \cite{CUDAProgrammingGuide}
for a detailed outlook.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/gpu-scaling.png}
\end{center}
\caption{Scheduling of blocks to streaming microprocessors---GPUs
    with more SMs are able to execute more blocks concurrently.
(\emph{CUDA Programming Guide ~\cite{CUDAProgrammingGuide}})}
\label{fig:gpu-scaling}
\end{figure}

@@

As mentioned in \ref{sec:tridiagonal-compact-schemes},
this work is concerned with
\emph{tridiagonal} compact finite differences
for their relative ease in evaluation.
The resulting tridiagonal system must be solved
repeatedly for all lines of grid points in the computational grid.
For a regular Cartesian grid (generally employed in DNS),
the coefficients of the tridiagonal systems
are the same for each grid line,
and only the right hand sides are different.
Thus, the compact finite difference evaluation
effectively requires the solution of a tridiagonal system
for several right hand sides.
In this section,
we describe the applicability of some algorithms
for solving this problem on the GPU.
We refer to ~\cite{chang2014guide} for
a more comprehensive list of algorithms,
along with details of their GPU implementation.

@@

Traditionally,
the Thomas algorithm is employed for solving
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
[Eq. (\ref{eqn:general-tridiagonal-system})].
The algorithm is derived
from the more general Gaussian (LU) algorithm
applied to tridiagonal matrices.
Implementations of this algorithm are relatively straightforward
and compact ~\cite{numericalrecipes}.
Only the non-zero diagonals of the tridiagonal matrix
are stored,
and the algorithm performs $2N$ steps
to solve the system.
It is also stable for diagonally dominant matrices.
The Thomas algorithm has an algorithmic complexity of $O(N)$,
and is the fastest serial solver for tridiagonal systems.
It is thus well suited to solution by a single CPU thread.
While the algorithm itself is sequential,
in a multithreaded environment,
the different CPU threads may be used to solve
the different independent tridiagonal systems simultaneously.

@@

The parallel cyclic reduction (PCR) algorithm
has only the forward reduction phase.
The first forward reduction step is applied to
the odd and even indexed equations seperately,
yielding two reduced systems
of size $n/2$.
Forward reduction applied to both of these
systems then yields \emph{four} reduced
systems of size $n/4$.
The process is repeated till
$n/2$ 2-by-2 systems are left,
all of which can be solved trivially.
The PCR algorithm requires half the number of steps
required by CR ($log_2(n)$),
but does significantly more computation per-step.
Further, unlike CR,
PCR can be implemented free of bank conflicts
~\cite{Zhang2010FTS}.

>>>>

    \item When the derivative is expressed as
        some combination of the function values
        on \emph{both} sides of $i$, i.e,
        at $\hdots$, $i-2$, $i-1$, $i$, $i-1$, $i-2$, $\hdots$,
        we refer to the approximation as a
        \emph{central} difference.
\end{enumerate}
%
The approximation of higher order derivatives generally requires
inclusion of a larger number of points in the
finite difference approximation
(referred to as the finite difference \emph{stencil}).
For a given order of derivative,
finite difference approximation of arbitrary stencil widths
may be derived,
with larger stencils associated with higher accuracy \cite{fornberg1988generation}.
In the evaluation of spatial derivatives,
the above schemes are referred to as \emph{explicit}
schemes, as the derivative at each point
can be expressed as some explicit combination of function values.

@@

Compact schemes express the derivative at a point $i$
in terms of
function values \emph{and derivatives}
at neighbouring points,
i.e., the derivative is expressed \emph{implicitly}.
For example,
if $f_i$ represents the value of
a uniformly sampled function evaluated at the $i$th sample point,
the first derivative $f^{\prime}_i$ can be approximated from
a relation of the form:
%
\begin{equation}
\begin{split}
    f_i^{\prime} + \
    \alpha(f^{\prime}_{i-1} + f^{\prime}_{i+1}) + \
    \beta(f^{\prime}_{i-2} + f^{\prime}_{i+2}) + \
    \hdots
    = 
    a\frac{f_{i+1} - f_{i-1}}{h} + \
    b\frac{f_{i+2} - f_{i-2}}{h} + \\
    c\frac{f_{i+3} - f_{i-3}}{h} + \
    \hdots
\end{split}
\label{eqn:general-compact}
\end{equation}
%
where $\alpha$, $\beta$, $a$, $b$, $c$, etc.,
are parameters that must satisfy certain constraints \cite{kennedy1994several,lele1992compact},
and $h$ is the space between two grid points.
We note that for equations of the form in Eq. (\ref{eqn:general-compact}),
the derivative at any points $i$ cannot be computed explicitly.
Instead, we must write similar equations for \emph{all} points $i$
in the range.
This results in a system of linear equations
with unknowns \{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
which may be represented by the matrix system $A\bm{x}=\bm{d}$,
where $\bm{x}$ is the vector
\{$f^{\prime}_1$, $f^{\prime}_2$, $\hdots$, $f^{\prime}_n$\},
$\bm{d}$ is a vector of right hand sides [Eq. (\ref{eqn:general-compact})],
and $A$ is, in general, a \emph{banded} matrix.

@@

The ability of a finite difference scheme
to accommodate large wavenumbers
is extremely relevant in
computational fluid dynamics applications
\cite{KravchenkoEffNumErr}.
In DNS, the scheme
\emph{must} capture the
rapidly varying characteristics of the flow
associated with the turbulence.
For this purpose,
higher order explicit schemes
may be considered.
While they are straightforward to compute,
they are associated with large stencil widths.
In most applications,
large stencil widths are undesirable.
This is because
the arithmetic intensity increases with stencil size, i.e.,
a larger number of computations must be performed per grid point.
Further,
the amount of boundary information that must be exchanged
between parallel processes increases,
which can seriously affect overall performance.

@@

One of the advantages of the compact finite difference approach
is that it accommodates \emph{non-periodic} boundary conditions.
This is in contrast to other methods used in DNS,
such as \emph{spectral} methods.
We note that the Eq. (\ref{eqn:general-compact})
cannot be applied near the boundary points.
At the boundaries,
\emph{non-centered} or \emph{one-sided}
finite difference approximations are required.
Some considerations are made in choosing
these approximations:
firstly, the bandwidth of the resulting banded matrix
must be preserved.
Secondly,
the width of the boundary stencils
must be lower than the interior stencils,
as higher order boundary stencils are unstable
\cite{kennedy1994several}.
In general, boundary equations for the first derivative
are of the following form \cite{lele1992compact}:

@@

Similar tridiagonal schemes are available
to evaluate higher order derivatives.
Compact schemes for
second, third and fourth derivatives
are derived in \cite{lele1992compact}.

@@

The evolution of graphics processing units (GPUs)
has traditionally been guided by the requirements
of graphics applications such as video gaming.
In general, these graphics computations involve
manipulations and calculations with pixel data.
Because these calculations are largely independent,
the GPU hardware is specialized
to perform several of them \emph{in parallel}.
Therefore, the
GPU may be regarded as a highly multithreaded processor.
Initially,
programming GPUs for scientific computation tasks
was a cumbersome process,
as the programming interfaces were designed for graphics computations,
and attempts to use them for other purposes were essentially ``hacks''.
Moreover, there was little support for double precision arithmetic,
which is critical for many applications.
But with the introduction of the
NVIDIA Compute Unified Device Architecture (CUDA)
parallel computing platform
and programming model,
CUDA-enabled GPUs have come into the
mainstream in high-performance computing.
A history of the development of GPUs
for general purpose computation,
and of CUDA specifically is available
in \cite{HwuBook}.
We provide the essential and relevant details about CUDA here,
and refer to the
CUDA Programming Guide \cite{CUDAProgrammingGuide}
for a detailed outlook.

@@

\begin{figure}
\begin{center}
\includegraphics[height=250pt]{img/gpu-scaling.png}
\end{center}
\caption{Scheduling of blocks to streaming microprocessors---GPUs
    with more SMs are able to execute more blocks concurrently.
(\emph{CUDA Programming Guide \cite{CUDAProgrammingGuide}})}
\label{fig:gpu-scaling}
\end{figure}

@@

As mentioned in \ref{sec:tridiagonal-compact-schemes},
this work is concerned with
\emph{tridiagonal} compact finite differences
for their relative ease in evaluation.
The resulting tridiagonal system must be solved
repeatedly for all lines of grid points in the computational grid.
For a regular Cartesian grid (generally employed in DNS),
the coefficients of the tridiagonal systems
are the same for each grid line,
and only the right hand sides are different.
Thus, the compact finite difference evaluation
effectively requires the solution of a tridiagonal system
for several right hand sides.
In this section,
we describe the applicability of some algorithms
for solving this problem on the GPU.
We refer to \cite{chang2014guide} for
a more comprehensive list of algorithms,
along with details of their GPU implementation.

@@

Traditionally,
the Thomas algorithm is employed for solving
general linear systems of the form $A\bm{x} = \bm{d}$,
where $A$ is a tridiagonal matrix with diagonals
$\bm{a}$, $\bm{b}$ and $\bm{c}$
[Eq. (\ref{eqn:general-tridiagonal-system})].
The algorithm is derived
from the more general Gaussian (LU) algorithm
applied to tridiagonal matrices.
Implementations of this algorithm are relatively straightforward
and compact \cite{numericalrecipes}.
Only the non-zero diagonals of the tridiagonal matrix
are stored,
and the algorithm performs $2N$ steps
to solve the system.
It is also stable for diagonally dominant matrices.
The Thomas algorithm has an algorithmic complexity of $O(N)$,
and is the fastest serial solver for tridiagonal systems.
It is thus well suited to solution by a single CPU thread.
While the algorithm itself is sequential,
in a multithreaded environment,
the different CPU threads may be used to solve
the different independent tridiagonal systems simultaneously.

@@

The parallel cyclic reduction (PCR) algorithm
has only the forward reduction phase.
The first forward reduction step is applied to
the odd and even indexed equations seperately,
yielding two reduced systems
of size $n/2$.
Forward reduction applied to both of these
systems then yields \emph{four} reduced
systems of size $n/4$.
The process is repeated till
$n/2$ 2-by-2 systems are left,
all of which can be solved trivially.
The PCR algorithm requires half the number of steps
required by CR ($log_2(n)$),
but does significantly more computation per-step.
Further, unlike CR,
PCR can be implemented free of bank conflicts
\cite{Zhang2010FTS}.

### 2015-11-18 thesis/proposed-tridiagonal-algorithm.tex
remove space after cite

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
~\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an approach for solving such
\emph{near-Toeplitz} tridiagonal systems efficiently on GPUs.

>>>>

\begin{equation} \label{eqn:toeplitz-matrix}
A = 
\begin{pmatrix}
     b_1 & c_1  \\
     a_0 & b_0  &  c_0  \\
         & a_0  &  b_0 &  c_0  \\
         &      &  a_0 &  b_0 &  c    \\
         &      &      &      &  \ddots \\
         &      &      &      &     &  \ddots  \\
         &      &      &      &     &  a_n  &  b_n
\end{pmatrix}
\end{equation}
%
where, in general
\begin{align*}
    a_0 &\neq a_n  \\
    b_1 &\neq b_0 \neq b_n \\
    c_1 &\neq c_0 
\end{align*}
%
We refer to matrices with this specific structure as
\emph{near-Toeplitz tridiagonal matrices},
and the corresponding linear systems as
\emph{near-Toeplitz tridiagonal systems}.
These matrices appear in a wide range of applications
\cite{sun1995application}
such as alternating direction implicit methods,
line relaxation methods,
and numerical solutions to one-dimensional differential equations.
We describe an approach for solving such
\emph{near-Toeplitz} tridiagonal systems efficiently on GPUs.
